Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-03-07 numerics_paper/version2/numerical_chj.tex
two versions of Ri and Di

\newcommand{\out}{\textup{out}}
\newcommand{\jet}{{\textup{jet}}}
\newcommand{\W}{{\mathcal W}}
\newcommand{\A}{{\mathcal A}}
\newcommand{\apdq}{\A^+ \!\!{\Delta} Q}
\newcommand{\amdq}{\A^- \!\!{\Delta} Q}
\newcommand{\Ft}{\tilde{F}}
\newcommand{\imh}{{i-1/2}}
\newcommand{\iph}{{i+1/2}}
\newcommand{\bff}{{f}}
\newcommand{\bfr}{{r}}
\newcommand{\bfF}{{F}}
\newcommand{\bfu}{{u}}
\newcommand{\entvar}{\eta'}
\newcommand{\bfq}{{Q}}
\newcommand{\bfx}{{x}}
\newcommand{\bfs}{{s}}
\newcommand{\bfn}{{n}}
\newcommand{\sgn}{{\rm sgn}}
\newcommand{\Rus}{{\rm Rus}}
\newcommand{\Roe}{{\rm Roe}}
\newcommand{\EV}{{\rm EV}}
\newcommand{\efp}{\psi}
\newcommand{\entflux}{g}
\newcommand{\Fr}{{\rm Fr}}

@@

\subsection{The entropy residual}\label{sec:entres}
Let $\entvar(q)$ and $\entflux(q)$ denote the entropy variable and entropy flux.
Recall that $q=[h,hu,hv]^T$. We use the total energy as entropy; that is, 
\begin{align*}
  \eta(q)=\frac{1}{2} gh^2 + \frac{(hu)^2+(hv)^2}{2h}, \qquad
  \entflux(q)=
    \frac{\eta(q)}{h}
    \begin{bmatrix} hu \\ hv \end{bmatrix}.
\end{align*}
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \entvar(q) \cdot \left[ \frac{\partial q}{\partial t} + \nabla\cdot \bff(q)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(q)}{\partial t} + \entvar(q) \cdot \nabla\cdot \bff(q)\right] d\bfx
\end{align}
as a measurement of the entropy production in cell $i$ (here $S_i=[x_\imh,x_\iph]$).
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(q)}{\partial t}d\bfx=-\int \nabla\cdot\entflux(q) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\entvar(q) \cdot \nabla\cdot \bff(q) - \nabla\cdot\entflux(q) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound on $R_i$ so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $q$ is smooth in $S_i$.
In our implementation, we use
\begin{subequations}\label{Ri_and_Di}
\begin{align}
  R_i 
  \approx 
  \left|\entvar(\bfq_i)\cdot \int_{S_i}\nabla\cdot \bff(q)d\bfx
  -\int_{S_i}\nabla\cdot\entflux(q) d\bfx\right| 
  = 
  \left|\entvar(\bfq_i)\cdot \int_{\partial S_i}\bfn_i(s) \cdot \bff(q) d\bfs
  -\int_{\partial S_i}\bfn_i(s)\cdot\entflux(q) d\bfs\right|, 
\end{align}
and 
\begin{align}
  D_i 
  = 
  \sum_{k=1}^{d+1}|\entvar_k(\bfq_i)|\left|\int_{\partial S_i}\left(\bfn_i(s)\cdot\bff(q)\right)_kd\bfs\right|
  +\left|\int_{\partial S_i}\bfn_i(s)\cdot\entflux(q) d\bfs\right|,
\end{align}
\end{subequations}
where $\bfn_i$ denotes the outward-pointing normal vector at the boundary of cell $i$, 
$(z)_k$ denotes the $k$-th component of $z\in\mathbb{R}^{d+1}$, 
and $d$ is the number of physical dimensions.  
In the integrands of \eqref{Ri_and_Di}, $q$ is the average of $Q$ 
from each side of the boundary $\partial S_i$. 
In \eqref{blended-flux-1} we need values of $\theta$ at the cell interfaces, for
which we use
$$
    \theta_\iph = \max(\theta_i, \theta_{i+1}).
$$
We expect that $\theta_\iph\approx 0 $ in smooth regions,
while $\theta_\iph\approx 1$ near shocks.
The value of $\theta_\iph$ controls whether the first-order flux \eqref{blended-flux-1}
behaves more like that of Rusanov or Roe.  Specifically, if
$\lambda_\iph^{\min}=0$ and $\theta_\iph=1$, the flux is equivalent to that of
Rusanov, while if $\lambda_\iph^{\min}=0$ and $\theta_\iph=0$, it is equivalent
to that of Roe.

>>>>

\newcommand{\out}{\textup{out}}
\newcommand{\jet}{{\textup{jet}}}
\newcommand{\W}{{\mathcal W}}
\newcommand{\A}{{\mathcal A}}
\newcommand{\apdq}{\A^+ \!\!{\Delta} Q}
\newcommand{\amdq}{\A^- \!\!{\Delta} Q}
\newcommand{\Ft}{\tilde{F}}
\newcommand{\imh}{{i-1/2}}
\newcommand{\iph}{{i+1/2}}
\newcommand{\bff}{{f}}
\newcommand{\bfr}{{r}}
\newcommand{\bfF}{{F}}
\newcommand{\bfu}{{u}}
\newcommand{\entvar}{\eta'}
\newcommand{\bfq}{{Q}}
\newcommand{\bfx}{{x}}
\newcommand{\bfs}{{s}}
\newcommand{\bfn}{{n}}
\newcommand{\f}{{\mathcal{F}}}
\newcommand{\sgn}{{\rm sgn}}
\newcommand{\Rus}{{\rm Rus}}
\newcommand{\Roe}{{\rm Roe}}
\newcommand{\EV}{{\rm EV}}
\newcommand{\efp}{\psi}
\newcommand{\entflux}{g}
\newcommand{\Fr}{{\rm Fr}}

@@

\subsection{The entropy residual}\label{sec:entres}
Let $\entvar(q)$ and $\entflux(q)$ denote the entropy variable and entropy flux.
Recall that $q=[h,hu,hv]^T$. We use the total energy as entropy; that is, 
\begin{align*}
  \eta(q)=\frac{1}{2} gh^2 + \frac{(hu)^2+(hv)^2}{2h}, \qquad
  \entflux(q)=
    \frac{\eta(q)}{h}
    \begin{bmatrix} hu \\ hv \end{bmatrix}.
\end{align*}
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \entvar(q) \cdot \left[ \frac{\partial q}{\partial t} + \nabla\cdot \bff(q)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(q)}{\partial t} + \entvar(q) \cdot \nabla\cdot \bff(q)\right] d\bfx
\end{align}
as a measurement of the entropy production in cell $i$ (here $S_i=[x_\imh,x_\iph]$).
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(q)}{\partial t}d\bfx=-\int \nabla\cdot\entflux(q) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\entvar(q) \cdot \nabla\cdot \bff(q) - \nabla\cdot\entflux(q) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound on $R_i$ so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $q$ is smooth in $S_i$.
In our implementation, we consider
{\color{blue}
\begin{align}
  R_i 
  &\approx 
  \left|\entvar(\bfq_i)\cdot \int_{S_i}\nabla\cdot \bff(q)d\bfx
  -\int_{S_i}\nabla\cdot\entflux(q) d\bfx\right| 
  = 
  \left|\entvar(\bfq_i)\cdot \int_{\partial S_i}\bfn_i(s) \cdot \bff(q) d\bfs
  -\int_{\partial S_i}\bfn_i(s)\cdot\entflux(q) d\bfs\right|,
\end{align}
where $\bfn_i$ denotes the unit vector normal to $\partial S_i$ pointing outward from cell $i$. 
To compute these integrals, we first split $\partial S_i$ into its faces $\{\f_i\in\partial S_i\}$.
Let $|\f_i|$ be the length of $\f_i$, $\bar Q_{\f_i}$ be the average of $Q$ 
from both sides of $\f_i$ and $\bfn_{\f_i}$ be the unit vector normal to $\f_i$ 
pointing outward from cell $i$.
Then, we use
\begin{subequations}\label{Ri_and_Di}
\begin{align}
  R_i = 
  \left|\sum_{\f_i\in\partial S_i} |\f_i|
  \left[\entvar(\bfq_i)\cdot \left(\bfn_{\f_i} \cdot \bff\left(\bar Q_{\f_i}\right)\right)-\bfn_{\f_i} \cdot\entflux \left(\bar Q_{\f_i}\right)\right] \right|,
\end{align}
and 
\begin{align}
  D_i 
  = 
  \sum_{k=1}^{d+1}\left|\entvar_k\left(\bfq_i\right)\right|
  \left|\sum_{\f_i\in\partial S_i}|\f_i|\left(\bfn_{\f_i}\cdot\bff\left(\bar Q_{\f_i}\right)\right)_k\right|
  +\left|\sum_{\f_i\in \partial S_i} |\f_i| \bfn_{\f_i}\cdot\entflux\left(\bar Q_{\f_i}\right) \right|,
\end{align}
\end{subequations}
where $(z)_k$ denotes the $k$-th component of $z\in\mathbb{R}^{d+1}$, 
and $d$ is the number of physical dimensions.}
{\color{red}
\begin{subequations}\label{Ri_and_Di_tmp}
\begin{align}
  R_i 
  \approx 
  \left|\entvar(\bfq_i)\cdot \int_{S_i}\nabla\cdot \bff(q)d\bfx
  -\int_{S_i}\nabla\cdot\entflux(q) d\bfx\right| 
  = 
  \left|\entvar(\bfq_i)\cdot \int_{\partial S_i}\bfn_i(s) \cdot \bff(q) d\bfs
  -\int_{\partial S_i}\bfn_i(s)\cdot\entflux(q) d\bfs\right|, 
\end{align}
and 
\begin{align}
  D_i 
  = 
  \sum_{k=1}^{d+1}|\entvar_k(\bfq_i)|\left|\int_{\partial S_i}\left(\bfn_i(s)\cdot\bff(q)\right)_kd\bfs\right|
  +\left|\int_{\partial S_i}\bfn_i(s)\cdot\entflux(q) d\bfs\right|,
\end{align}
\end{subequations}
where $\bfn_i$ denotes the unit vector normal to $\partial S_i$ pointing outward from cell $i$,
$(z)_k$ denotes the $k$-th component of $z\in\mathbb{R}^{d+1}$, 
and $d$ is the number of physical dimensions.  
In the integrands of \eqref{Ri_and_Di_tmp}, $q$ is the average of $Q$ 
from each side of the boundary $\partial S_i$.}
In \eqref{blended-flux-1} we need values of $\theta$ at the cell interfaces, for
which we use
$$
    \theta_\iph = \max(\theta_i, \theta_{i+1}).
$$
We expect that $\theta_\iph\approx 0 $ in smooth regions,
while $\theta_\iph\approx 1$ near shocks.
The value of $\theta_\iph$ controls whether the first-order flux \eqref{blended-flux-1}
behaves more like that of Rusanov or Roe.  Specifically, if
$\lambda_\iph^{\min}=0$ and $\theta_\iph=1$, the flux is equivalent to that of
Rusanov, while if $\lambda_\iph^{\min}=0$ and $\theta_\iph=0$, it is equivalent
to that of Roe.
